---
title: "Citigrapher #1: Google Maps API"
author: "OJ Watson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    css: knitr.css
    code_folding: show
    fig_caption: yes
    theme: readable
    keep_md: yes
    toc_float: yes

---

## Overview


1. User input 
2. API query
3. Storage & Extension

These 3 parts will be demonstrated in the following tutorial and will give an idea
of how a few packages can be used to tackle the problem. 

## 1. User input

The first thing that a user will want to do when using citigrapher is to input locations
that represent their starting and final destinations. These may represent their work and
home addresses respectively. We will want to be able to do this for a number of potential
people.

```{r User input, echo=FALSE, include=TRUE, message = FALSE, warning = FALSE}

request_user_address <- function(){

## Ask user for start address
start.address <- readline(prompt="Enter a start location: ")

## Ask user for final address
final.address <- readline(prompt="Enter a final location: ")

## And let's returnn this for further use
return(list("Start"=start.address, "Final"=final.address))

}

```

We can now use this within a larger function to get everyone's addresses:

```{r User inputs, echo=FALSE, include=TRUE, message = FALSE, warning = FALSE}

request_all_user_addresses <- function(){

## Ask user for number of people
num.people <- readline(prompt="How many people are we connecting: ")

## Preinitialise results
res <- list()
length(res) <- num.people

## Query this many times
for(i in 1:num.people){
  ## give the user an idea of who the info is for
  message(paste0("For person ",i))
  res[[i]]$Location <- request_user_address() 
}

## And let's returnn this for further use
return(res)

}

```

The above will clearly be insufficient/annoying when there are lots of people, so in a future meetup
we will be exploring how to best carry out local storage for a user so that they only ever have to enter
information for a particular person once. I.e. if you have already entered your friend Bob's work and 
home addresses you don't want to do that every time, but simply ask the user whether they want to use
Bob's default addresses. 

We will now pretend that we saved this output to file so we can use it in the next step, so we will 
now import it here:

```{r Load addresses, include=TRUE, message = FALSE, warning = FALSE}

user_addresses <- read.delim("inst/extdata/user_addresses.txt", 
                             sep="\t",
                             stringsAsFactors = FALSE)

## turn it into a list output like above
res <- list()
length(res) <- length(user_addresses)
for(i in 1:length(user_addresses)){
  res[[i]]$Location$Start <- user_addresses$Start[i]
  res[[i]]$Location$End <- user_addresses$End[i]
}

```

## 2. API query

With the addresses we will want to isolate which tube stops are best for each person's start and
end destination. To do this we will use the igraph tube map object that is saved within the citigrapher repo,
along with the ggmap package which is superb. First we will need to convert our addresses to lat long

```{r Google lat long query, include=TRUE, message = FALSE, warning = FALSE}

## first let's load necessary packages
require(ggmap)
require(igraph)

## Then let's build a function to query google for the lat long of our postcodes
## There is lot's of extra info in here which is not needed now, but will be useful in 
## later meetups
get_geo_details <- function(location){   
  
  # use the gecode function to query google servers
  geo_reply = ggmap::geocode(location, output='all', messaging=TRUE, override_limit=TRUE)
  
  # now extract the bits that we need from the returned list
  answer <- data.frame(lat=NA, long=NA, accuracy=NA, formatted_address=NA, address_type=NA, status=NA)
  answer$status <- geo_reply$status
  
  # return Na's if we didn't get a match:
  if (geo_reply$status != "OK"){
    return(answer)
  }   
  
  # else, extract what we need from the Google server reply into a dataframe:
  answer$lat <- geo_reply$results[[1]]$geometry$location$lat
  answer$long <- geo_reply$results[[1]]$geometry$location$lng   
  
  # if the reply contained the information we want
  if (length(geo_reply$results[[1]]$types) > 0){
    answer$accuracy <- geo_reply$results[[1]]$types[[1]]
  }
  
  # paste together the answer
  answer$address_type <- paste(geo_reply$results[[1]]$types, collapse=',')
  answer$formatted_address <- geo_reply$results[[1]]$formatted_address
  
  return(answer)
}

```

We can now use this within a larger function to work out the closest 3 tubes to 
our requested addresses.

```{r Closest tubes, include=TRUE, message = FALSE, warning = FALSE, cache=TRUE}

## First load the igraph tube map
tube.map <- readRDS("inst/extdata/tubemap.rds")

## Quick view of the igraph object
tube.map

## From this object we can then query our lat/longs against the station

## First let's get the lat/long for our addresses
lat.long.starts <- lapply(res,function(x){return(get_geo_details(x$Location$Start))})
lat.long.ends <- lapply(res,function(x){return(get_geo_details(x$Location$End))})

## For now let's just keep the lat and long componenets in our res

for(i in 1:length(res)){
  
  res[[i]]$Location$StartLat <- lat.long.starts[[i]]$lat
  res[[i]]$Location$StartLong <- lat.long.starts[[i]]$long
  res[[i]]$Location$EndLat <- lat.long.ends[[i]]$lat
  res[[i]]$Location$EndLong <- lat.long.ends[[i]]$long
  
}


## We will then need to create a function to take the lat.long.starts and ends
## and find the nearest tube
get_closest_tubes <- function(lat, long, tube.map, num.tubes){
  
# first create matrix of tube lat longs  
tube.lat.long.mat <- cbind(get.vertex.attribute(tube.map,"latitude"),get.vertex.attribute(tube.map,"longitude"))

# create vector of distances as crow flies
distances <- sqrt(abs(tube.lat.long.mat[,1] - lat)^2 + abs(tube.lat.long.mat[,2] - long)^2)

# find the cloesest number of tubes requested
closests <- head(sort(distances),num.tubes)

# return the name of the tube 
return(get.vertex.attribute(tube.map,"name")[match(closests,distances)])

}
```

We can now now use the function to find the closest 3 tubes for each user's start and end. (You
may have also noticed that it makes good sense to probably move the tube.lat.long.mat outside the
function but for the sake of clarity we'll leave it in.)

```{r Add closeset tubes, include=TRUE, message = FALSE, warning = FALSE, cache=TRUE}

# By simply looping over the function we can add the required tubes
for(i in 1:length(res)){
  
  res[[i]]$StartTubes <- get_closest_tubes(lat = res[[i]]$Location$StartLat, 
                                            long = res[[i]]$Location$StartLong,
                                            tube.map = tube.map,
                                            num.tubes = 3)
  
  res[[i]]$EndTubes <- get_closest_tubes(lat = res[[i]]$Location$EndLat,
                                            long = res[[i]]$Location$EndLong,
                                            tube.map = tube.map,
                                            num.tubes = 3)
  
}

## quick look at what we now have
res

```

At this point we now know the closest geographical tubes, but we don't know the actual 
walking distance to the tubes. This will ultimately be very important when trying to 
find the optimum tube in later weeks. To address this we will now query the google
maps API to find the walking distances for each person/tube combination.

```{r Query google maps api, include=TRUE, message = FALSE, warning = FALSE, cache=TRUE}

## function to add travel times to our results object thus far
add_tube_distance_times <- function(res, API_KEY){

# Loop over each person
for(i in 1:length(res)){

# First find the lat/long for the start and end origin, i.e. the user provided postcode lat/longs
start.origin <- paste(res[[i]]$Location$StartLat,",",res[[i]]$Location$StartLong,sep="")
end.origin <- paste(res[[i]]$Location$EndLat,",",res[[i]]$Location$EndLong,sep="")

# Find which position in the tube.map the start and end tubes are
tube.matches.start <- match(res[[i]]$StartTubes,get.vertex.attribute(tube.map,"name"))
tube.matches.end <- match(res[[i]]$EndTubes,get.vertex.attribute(tube.map,"name"))

# Fetch the lat/longs for these tubes
start.destination.lats <- get.vertex.attribute(tube.map,"latitude")[tube.matches.start]
start.destination.longs <- get.vertex.attribute(tube.map,"longitude")[tube.matches.start]
end.destination.lats <- get.vertex.attribute(tube.map,"latitude")[tube.matches.end]
end.destination.longs <- get.vertex.attribute(tube.map,"longitude")[tube.matches.end]

# paste the lat/longs together ready for url_query creation
start.destinations <- paste(start.destination.lats,start.destination.longs,sep=",",collapse="|")
end.destinations <- paste(end.destination.lats,end.destination.longs,sep=",",collapse="|")

# url query for start
url.query.start <- paste0("https://maps.googleapis.com/maps/api/distancematrix/json?units=imperial",
                    "&origins=",start.origin,
                    "&destinations=",start.destinations,
                    "&mode=walking",
                    "&key=",API_KEY)

# url query for end
url.query.end <- paste0("https://maps.googleapis.com/maps/api/distancematrix/json?units=imperial",
                    "&origins=",end.origin,
                    "&destinations=",end.destinations,
                    "&mode=walking",
                    "&key=",API_KEY)

# Now query google maps api distance matrix
query.results.start <-jsonlite::fromJSON(url.query.start, simplifyDataFrame = FALSE)
query.results.end <-jsonlite::fromJSON(url.query.end, simplifyDataFrame = FALSE)

# Add the results
res[[i]]$StartTubesDistanceTimes <- unlist(lapply(query.results.start$rows[[1]]$elements,function(x){return(x$duration$value)}))
res[[i]]$EndTubesDistanceTimes <- unlist(lapply(query.results.end$rows[[1]]$elements,function(x){return(x$duration$value)}))

}

return(res)
  
}

## So let's now use this to add the travel times
res <- add_tube_distance_times(res,"AIzaSyAcqeRRuaTpG-YGhyCuw6ijv0QwCBBYw5c")

## quickly view what the res object looks like now
res
```

## 3. Storage and Extension

The last section is shorter and more open ended for everyone to start thinking
about how might be best to store data as citigrapher grows, and also how best 
to improve on this section. 

Firstly, at the moment the storage is a nested list object that was designed on the fly
while writing this. Over the course of citigrapher we will possibly find that this
is not the most useful storage format, especially if we want it to be cross programming
platform. So we might want to start thinking about json storage as one example.

```{r json conversion, message=FALSE, warning=FALSE, include=TRUE, results="hide"}

## Json conversion and writing
res_json <- jsonlite::toJSON(res)
jsonlite::write_json(res_json,"inst/extdata/user_addresses_plus.json")

```

We also might want to instantly select the current user's location, to save them the
hassle. There are a number of ways to do this, but one could be as simple as pinging 
any number of find my location websites, e.g. https://mycurrentlocation.net/.  

---

## Summary

Hopefully the above tutorial has shown a framework in R of how we can start working out 
how long it will take a set of individuals to get to their nearest tubes. This will then form
one section of a sum that will then workout each person's travel times to a particular tube
stop. This total sum will then represent the "score" for a particular tube stop, and through
a heuristic search method we can start to see how we can find optimum tubes. This will involve
being more familiar with utilising networks, which will likely be the topic for next time.
